Individual and Differential Harm in Redistricting

### 2025-07-29 paper/harm.qmd
trimming

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp. 72--73], is perhaps the most consistently visible manifestation of this tendency. But gerrymandering takes other forms as well. In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial]. At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the districting plan may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process]. The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].

The academic study of gerrymandering has unfortunately not reflected this diversity.
The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020]. A large number of formulas for quantifying or identifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18]. The study of racial gerrymandering has been largely shaped by the U.S. legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA). This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological]. Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

The dominance of the partisan framework for analyzing redistricting poses a problem even in cases of pure partisan gerrymandering.
Existing tools, by focusing only on the effect of district lines on parties, are unable to examine the effect of manipulated lines on other groups.
A redistricting plan that diminishes Democratic voting power will naturally have an effect on the relative voting power of White and Black voters, poor and wealthy voters, young and old voters, and urban or rural voters.
These effects on individuals will be missed by any framework that only studies parties.

This paper addresses these limitations with a new framework of *individual and differential harm* for the unified study of all forms of gerrymandering.
Fundamentally, gerrymandering involves the concentration or dilution of voting power to favor or disfavor a particular group.
Our proposed framework focuses directly on measuring this concentration and dilution, and in doing so avoids the pitfalls of party-centric approaches.

The harm framework combines two key ideas, which are motivated at greater length in Section @sec-motivation.
First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of a districting plan.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

The combination of individual voter choices and counterfactual districting plans defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
Section @sec-motivation illustrates this with a simple example, and the general definitions are formalized in Section @sec-indiv-harm.

A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022]. This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry]. The challenge of estimating harm and differential harm from observed data is discussed in Section @sec-estimating. Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section @sec-connections.

Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
A brief study of partisan gerrymandering of New Jersey congressional districts in Section @sec-nj shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
Section @sec-alabama, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
Section @sec-boston demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
Section @sec-discussion concludes.

@@

The core idea motivating individual and differential harm is a simple one.
Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure @fig-minissouri.
The state's total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two medium-sized cities and a substantial rural population with an overall partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.

@@

Now suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
In contrast, a more proportional plan @fig-minissouri(b) elects two Republicans and one Democrat.

Intuitively, plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
These voters are highlighted in Figure @fig-minissouri(c).
For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district in both the enacted and counterfactual plan.

@@

We need not limit our consideration to partisanship.
We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
In other words, square voters are 20 percentage points more likely to be harmed than circle voters.
While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b), and it does so on the same scale as differential partisan harm.

@@

Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
In the worst cases, this can lead to directionally wrong conclusions, as we see in the next section.

@@

```{r fig-boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
#| fig.cap: |
#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
#|      ellipses (left), and geographic variation in factor values by race (right).
#|      Neighborhood boundaries overlaid in black.
#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
include_graphics("figures/boston_factors.pdf")
```

@@

>>>>

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, going back to Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp. 72--73], is perhaps the most consistently visible manifestation of this tendency.
But gerrymandering takes other forms as well.
In the United States, states historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
At the local level, where most redistricting occurs, partisanship is less salient, and manipulation may be more subtle or may favor a narrower group [@cain2002mapmaking; @halberstam2012process].
The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].

The academic study of gerrymandering has not reflected this diversity.
Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

The dominance of partisan frameworks limits redistricting analysis, even in purely partisan cases.
By focusing only on party effects, existing tools overlook how partisan lines also have an effect of relative voting power by race, wealth, age, and urbanicity.

This paper addresses these limitations with a new framework of *individual and differential harm*.
Our proposed framework focuses on measuring the concentration or dilution of voting power to favor or disfavor a particular group.
This avoids the pitfalls of party-centric approaches.

The harm framework combines two key ideas, further motivated in Section @sec-motivation.
First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].

Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
Many factors influence the performance and qualities of a districting plan.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw even one reliable Republican seat, despite 32% of voters being Republican [@duch:18].
The 9-0 Democratic composition reflects the homogeneity of the state's political geography, not partisan gerrymandering.
Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
Many metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

The combination of individual voter choices and counterfactual districting plans defines *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
By examining how harm varies across partisan, racial, or ideological lines, we link it back to traditional gerrymandering.
Section @sec-motivation gives a simple example Section @sec-indiv-harm formalizes it.

A key feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
The challenge of estimating harm and differential harm from observed data is discussed in Section @sec-estimating.
Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].

Our proposed harm measures improve on exististing racial and partisan gerrymandering analytical approaches.
The harm approach corresponds well to measuring the Gingles criteria, can directly use data from ecological inference, and handles complex cases like multi-minority coalitions [@stephanopoulos2014south]
Further, we explicitly model the counterfactuals and their feasibility, allowing us to improve on existing issues with other measures such as partisan bias [@king1987democratic], efficiency gap [@stephanopoulos2015], and declination [@warrington18].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section @sec-connections.

Harm and differential harm offer practical value across diverse contexts.
Section @sec-nj analyzes New Jersey congressional districts, showing how traditional metrics can misidentify both the magnitude and direction of partisan gerrymanders.
Section @sec-alabama uses differential harm to disentangle racial and partisan effects in Alabama's districting.
Section @sec-boston applies the framework to local redistricting, quantifying the benefits of race-aware boundaries for Black voters in Boston.
Section @sec-discussion concludes.

@@

Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure @fig-minissouri.
Its total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two cities and a substantial rural population with a partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation (red or blue), each voter also belongs to a racial group (squares or circles).

@@

Suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
Though Republicans win the popular vote, Democrats win two of three seats.
In contrast, @fig-minissouri(b) is more proportional, electing two Republicans and one Democrat.

Intuitively, plan (a) unfairly packs Republicans into a rural district.
We can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
These voters are highlighted in Figure @fig-minissouri(c).
For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district under both plans.

@@

We need not limit our consideration to partisanship.
We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$.
This means square voters are 20 percentage points more likely to be harmed than circle voters.
While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or aimed to maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.

@@

@@

```{r fig-boston-factors, fig.width=7, fig.height=3.5}
#| fig.cap: |
#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
#|      ellipses (left), and geographic variation in factor values by race (right).
#|      Neighborhood boundaries overlaid in black.
#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
include_graphics("figures/boston_factors.pdf")
```

@@

Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
In the worst cases, this can lead to directionally wrong conclusions.
